---
title: "SpaMTP: Refinning Metabolite Annotations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Annotation_Refinement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "##",
  fig.width = 7,       # smaller inches
  fig.height = 4,      # proportionate aspect ratio
  dpi = 72,
  fig.align = 'center',
  dev.args = list(png = list(compression = 9))  # max compression
)
```


# Refinning m/z Metabolite Annotations with SpaMTP

This tutorial highlights how to use *SpaMTP* to refine m/z annotated with multiple metabolites. *SpaMTP* implements this in 4 different ways, with these functions:

1) CalculateAnnotationStatistics
2) RefineLipids
3) Pseudo_msms
4) Compare_msms

We will visit each in detail below.

Author: Andrew Causer

<br>


## Install and Import *R* Libraries

First we need to import the required libraries for this analysis.

```{r load_packages, warning=FALSE, message=FALSE}
## Install SpaMTP if not previously installed
if (!require("SpaMTP"))
    devtools::install_github("GenomicsMachineLearning/SpaMTP")

#General Libraries
library(SpaMTP)
library(Cardinal)
library(Seurat)
library(dplyr)
library(tidyr)

#For plotting + DE plots
library(ggplot2)
library(EnhancedVolcano)
library(viridis)
```



## 1) Pathway-Based Refinement: `CalculateAnnotationStatistics`

This function uses correlations in pathway expression (Metabolite only, Gene only or Multi Omic based pathway information) to refine metabolite annotation. Specifically, for each annotated metabolite, the expression profiles for all known pathways associated with this metabolite are calculated. Based on spatial colocalisation between relative pathways and the given m/z mass, each metabolite is then ranked based on Pearson correlation values and the number of significant pathways associated with that metabolite.

Below we will use a [public mouse liver dataset with spotted chemicals standards](https://zenodo.org/records/17289187) to demonstrate this:

```{r, include=FALSE}
options(timeout = 600)
```

```{r warning=FALSE, message=FALSE}
spotted <- readRDS(url("https://zenodo.org/records/17289187/files/spotted_Annotated_H_Cl_Adducts.RDS?download=1"))
```

```{r}
spotted
```


Here we have a SpaMTP Seurat object that is already processed, with m/z annotated. Meaning we have 1303 annotated m/z values. This dataset includes certain chemical compounds which were 'spiked-in' at known locations. We will plot these below: 

```{r fig.width=7, fig.height=4, dpi = 72, warning=FALSE, message=FALSE}
metabolite_colors <- list("ATP" = "azure","ADP" = "green","Adenosine" = "yellow", "Guanine" = "red","Glutamic Acid" = "turquoise1", "Glutamine" = "chocolate1","Glutathione" = "deeppink","liver" = "black")

ImageDimPlot(spotted, group.by = "clusters", cols = metabolite_colors, dark.background = F)
```

The plot highlights the location of each known metabolite. However, due to the inherent issue of mass spectrometry imaging data one m/z mass may be associated with multiple possible metabolites. For example *adenosine triphosphate* or ATP was spiked-in to the second set of wells within the liver (green). When we look at the annotation of ATP, we can see for `m/z-505.988470137` there are 4 different annotations, one of which is ATP. 

```{r, eval=FALSE, warning=FALSE, message=FALSE}
SearchAnnotations(spotted, metabolite = "Adenosine triphosphate", search.exact = T)
```

```{r, echo=FALSE}
df <- SearchAnnotations(spotted, metabolite = "Adenosine triphosphate", search.exact = T)
knitr::kable(df, format = "html", table.attr = "class='dataframe'", align = "c")
```


Plotting this m/z value spatially, it localises completely within the ATP spiked-in region. 

```{r, fig.width=10, fig.height=5, warning=FALSE, message=FALSE}
ImageDimPlot(spotted, group.by = "clusters", cols = setNames(
  ifelse(names(metabolite_colors) == "ATP", "azure", "darkgrey"),names(metabolite_colors)
))+ ggtitle("ATP")| ImageMZPlot(spotted, mzs = 505.9885)
```

This confirms that the correct annotation of `mz-505.988470137` is only 'ATP'. This is the case with all other spiked-in chemicals. To refine the annotations based on pathway correlation we can run SpaMTP's `CalculateAnnotationStatistics`:

```{r, eval=FALSE}
## Step 1: Create a pathway assay - This can be done using only metabolites, or can be merged to include genes if applicable
spotted <- CreatePathwayAssay(spotted, analyte_type = "metabolites", assay = "Spatial", new_assay = "SPM_pathway", verbose = T)

## Step 2: Format Pathway Assay
DefaultAssay(spotted) <- "SPM_pathway"
spotted <- NormalizeData(spotted)
spotted <- ScaleData(spotted)

## Step 3: Run CalculateAnnotationStatistics
output <- CalculateAnnotationStatistics(data = spotted, mz.assay = "Spatial",pathway.assay="SPM_pathway", mz.slot = "counts", return.top = FALSE)
```

After running `CalculateAnnotationStatistics` the output will be a dataframe containing the most likely annotation for each m/z value based on correlations with pathway expression. Lets subset this output to look at our spiked-in metabolites:

```{r, include=FALSE, warning=FALSE, message=FALSE}
output <- readRDS(url("https://zenodo.org/records/17352749/files/annotation_zscore_output.RDS?download=1"))
```

```{r, eval=FALSE, warning=FALSE, message=FALSE}
output[c(1162,1053,501,626,126,117,251,113,248,637,638,835,836),]
```

<div style="text-align: center;">

| n_sig_path | max_cor | z_score | pval | pval_adj | original_annotations |
| :---------: | :------: | :------: | :------: | :------: | :------ |
| 1162 | 505.9885 | mz-505.988470137 | 0.0416385 | 0.1249154 | Adenosine triphosphate; dGTP; 2-hydroxy-dATP; [[(2R,5R)-5-(6-Aminopurin-9-yl)-3,4-dihydroxyoxolan-2-yl]methoxy-hydroxyphosphoryl] phosphono hydrogen phosphate |
| 1053 | 426.0221 | mz-426.022139245 | 0.0055132 | 0.0275662 | Adenosine 3’,5’-diphosphate; dGDP; ADP; [(2S,3R,4R,5R)-5-(6-Aminopurin-9-yl)-3,4-dihydroxyoxolan-2-yl]methyl phosphono hydrogen phosphate; Zidovudine diphosphate |
| 501 | 266.0895 | mz-266.089477461 | 0.0000862 | 0.0006897 | 3’-Azido-3’-deoxythymidine, 98%; 8-Oxo-2’-deoxyadenosine; Adenosine; Deoxyguanosine; Vidarabine; Zidovudine; 9-Arabinofuranosyladenine; Formycin A |
| 626 | 302.0662 | mz-302.066155193 | 0.0000192 | 0.0001726 | Hpmpa; 3’-Azido-3’-deoxythymidine, 98%; 8-Oxo-2’-deoxyadenosine; Adenosine; Deoxyguanosine; Vidarabine; Zidovudine; 9-Arabinofuranosyladenine; Formycin A |
| 126 | 150.0421 | mz-150.042133345 | 0.0013499 | 0.0053996 | 8-Oxo-7,8-dihydrodeoxyguanine; Guanine; 2-Hydroxyadenine; 8-Hydroxyadenine |
| 117 | 146.0459 | mz-146.045881333 | 0.0000008 | 0.0000099 | N-lactoyl-Glycine; 5-Hydroxy-4-oxo-L-norvaline; Alanine acetate; N-Methyl-DL-aspartic acid; L-Glutamic acid; N-Methyl-D-aspartic acid; N-Acetylserine; O-Acetylserine; D-Glutamic acid; L-4-Hydroxyglutamate semialdehyde; 3-(Carboxymethylamino)propanoic acid; DL-Glutamate |
| 251 | 182.0226 | mz-182.022559065 | 0.0000000 | 0.0000004 | Fosmidomycin; 2-Amino-4-phosphonobutyric acid; N-lactoyl-Glycine; 5-Hydroxy-4-oxo-L-norvaline; Alanine acetate; N-Methyl-DL-aspartic acid; L-Glutamic acid; N-Methyl-D-aspartic acid; N-Acetylserine; O-Acetylserine; D-Glutamic acid; L-4-Hydroxyglutamate semialdehyde; 3-(Carboxymethylamino)propanoic acid; DL-Glutamate |
| 113 | 145.0619 | mz-145.061865745 | 0.0000203 | 0.0001626 | Alanylglycine; Cyclic Urea; Glycyl-D-Alanine; Glycylsarcosine; D-Alanyl glycine; L-Glutamine; Ureidoisobutyric acid; D-Glutamine; 3-ureido-isobutyrate |
| 248 | 181.0385 | mz-181.038543477 | 0.0000184 | 0.0001472 | Alanylglycine; Cyclic Urea; Glycyl-D-Alanine; Glycylsarcosine; D-Alanyl glycine; L-Glutamine; Ureidoisobutyric acid; D-Glutamine; 3-ureido-isobutyrate |
| 637 | 306.0765 | mz-306.076529999 | 0.0000223 | 0.0001337 | Glutathione; Glutathionate(1-); Methyl 2-((3,4-dihydro-3,4-dioxo-1-naphthalenyl)amino)benzoate; Hallacridone; Aristolodione; zaprinast |
| 638 | 306.0772 | mz-306.077181461 | 0.0000223 | 0.0001337 | Glutathione; Glutathionate(1-); Methyl 2-((3,4-dihydro-3,4-dioxo-1-naphthalenyl)amino)benzoate; Hallacridone; Aristolodione; zaprinast |
| 835 | 342.0532 | mz-342.053207731 | 0.0000223 | 0.0001337 | O-Desmethylindomethacin; Glutathione; Glutathionate(1-); Methyl 2-((3,4-dihydro-3,4-dioxo-1-naphthalenyl)amino)benzoate; Hallacridone; Aristolodione |
| 836 | 342.0539 | mz-342.053859193 | 0.0000223 | 0.0001337 | O-Desmethylindomethacin; Glutathione; Glutathionate(1-); Methyl 2-((3,4-dihydro-3,4-dioxo-1-naphthalenyl)amino)benzoate; Hallacridone; Aristolodione |

</div>


We can see for each spiked-in metabolite the predicted metabolite with the highest significance in each case corresponded to the spike-in. This method gives the user higher confidence that the annotations of m/z value are correct and biologically significant. Note: For large datasets this function can take a while to run, the function `CalculateSingleAnnotationStatistics` can be used instead to calculate refined annotations for a single m/z value. Please visit the [documentation](https://genomicsmachinelearning.github.io/SpaMTP/reference/CalculateSingleAnnotationStatistics.html) for more information.


## 2) Lipid Nomenchlature Refinement 

As described previously in the [`Spatial Metabolomics Analysis`](https://genomicsmachinelearning.github.io/SpaMTP/articles/Mouse_Urinary_Bladder.html) tutorial, because many different kinds of lipids can share the same molecular weight, this may results in some m/z values possessing many possible annotations. The SpaMTP function `RefineLipids` can be used to simplify the lipid nomenclature into common lipid categories and classes, giving more simplified annotations to a level that most SM datasets are capable of identifying. 

```{r, message=FALSE, warning=FALSE, results='hide'}
### Runs lipid nomenclature simplification on annotations
refined_lipid_annotations <- RefineLipids(spotted@assays$Spatial@meta.data, annotation.column = "all_IsomerNames", lipid_info = "simple")
```

We can count the number of metabolites assigned for each m/z before and after refining the lipid annotations.
```{r, message=FALSE, warning=FALSE}
## Count original number of annotations per m/z 
refined_lipid_annotations <- refined_lipid_annotations %>%
  mutate(base_count = sapply(strsplit(all_IsomerNames, ";\\s*"), length))

## Count number of annotations per m/z following lipid simplification
refined_lipid_annotations <- refined_lipid_annotations %>%
  mutate(
    refined_count = ifelse(
      is.na(Lipid.Maps.Category),
      base_count,
      sapply(strsplit(Lipid.Maps.Category, ";\\s*"), length)
    )
  )
```

The maximum reduction in annotations was from 161 to 1. This is quite a significant simplification which can help in processing large datasets. Lets visualise the difference between each m/z value below: 

```{r, fig.width=7, fig.height=7, dpi = 72, warning=FALSE, message=FALSE}
# reshape to long format
df_long <- refined_lipid_annotations %>%
  select(raw_mz, refined_count, base_count) %>%
  pivot_longer(cols = c(refined_count, base_count),
               names_to = "Type", values_to = "Count")

# plot
ggplot(df_long, aes(x = raw_mz, y = Count, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, size = 1.2) +
  scale_color_manual(values = c("refined_count" = "red",
                                "base_count" = "blue"),
                     labels = c("Base annotations", "Refined annotations")) +
  labs(
    x = expression(italic("m/z")),
    y = "Number of annotations",
    color = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.title = element_text(face = "bold")
  )
```

We can see the larger m/z values, which correspond mostly to lipids, display the biggest differences following annotation refinement. 


## 3) Pseudo MS/MS-Based Refinement

To come!

## 4) Refinement with Paired Targeted Metabolic Data

To come!





