---
title: "SpaMTP: Refining Metabolite Annotations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SpaMTP: Refining Metabolite Annotations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
download_helper_candidates <- c(
  file.path("vignettes", "vignette_download_utils.R"),
  "vignette_download_utils.R",
  file.path("..", "vignettes", "vignette_download_utils.R")
)
download_helper <- download_helper_candidates[
  file.exists(download_helper_candidates)
][[1]]
sys.source(download_helper, envir = knitr::knit_global())

knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "##",
  fig.width = 7,       # smaller inches
  fig.height = 4,      # proportionate aspect ratio
  dpi = 72,
  fig.align = 'center',
  dev.args = list(png = list(compression = 9))  # max compression
)
```


# Refining m/z Metabolite Annotations with SpaMTP

This tutorial highlights complementary ways to prioritise multiple candidate
metabolites assigned to the same observed m/z. The first two approaches are
implemented directly in *SpaMTP*. A custom matrix-adduct search can then extend
the candidate space for known acquisition chemistry. Pseudo-MS/MS and paired
targeted metabolomics provide additional independent evidence:

1. Pathway-based prioritisation with `CalculateAnnotationStatistics`
2. Lipid nomenclature simplification with `RefineLipids`
3. Custom matrix-adduct design and single-peak candidate comparison
4. Pseudo-MS/MS evidence from an independent MS1-ID/GNPS spectral database
5. Comparison with paired targeted metabolomics data

These approaches provide complementary evidence. They should prioritise or
support accurate-mass candidates rather than silently replace the original
annotations.

Author: Andrew Causer

<br>


## Install and Import *R* Libraries

First we need to import the required libraries for this analysis.

```{r load_packages, warning=FALSE, message=FALSE}
## Install SpaMTP if not previously installed
if (!require("SpaMTP"))
    devtools::install_github("GenomicsMachineLearning/SpaMTP")

#General Libraries
library(SpaMTP)
library(Cardinal)
library(Seurat)
library(dplyr)
library(tidyr)

#For plotting + DE plots
library(ggplot2)
library(EnhancedVolcano)
library(viridis)
```



## 1) Pathway-Based Refinement: `CalculateAnnotationStatistics`

This function uses correlations in pathway expression (Metabolite only, Gene only or Multi Omic based pathway information) to refine metabolite annotation. Specifically, for each annotated metabolite, the expression profiles for all known pathways associated with this metabolite are calculated. Based on spatial colocalisation between relative pathways and the given m/z mass, each metabolite is then ranked based on Pearson correlation values and the number of significant pathways associated with that metabolite.

Below we will use a [public mouse liver dataset with spotted chemicals standards](https://zenodo.org/records/17289187) to demonstrate this:

```{r warning=FALSE, message=FALSE}
spotted_file <- resolve_spamtp_vignette_file(
  article = "Annotation_Refinement",
  filename = "spotted_Annotated_H_Cl_Adducts.RDS",
  url = "https://zenodo.org/api/records/17289187/files/spotted_Annotated_H_Cl_Adducts.RDS/content",
  expected_md5 = "d7b46c8d71b7a3eb95a01a210e010c40"
)
spotted <- readRDS(spotted_file)
```

```{r}
spotted
```


Here we have a SpaMTP Seurat object that is already processed, with m/z annotated. Meaning we have 1303 annotated m/z values. This dataset includes certain chemical compounds which were 'spiked-in' at known locations. We will plot these below: 

```{r fig.width=7, fig.height=4, dpi = 72, warning=FALSE, message=FALSE}
metabolite_colors <- list("ATP" = "azure","ADP" = "green","Adenosine" = "yellow", "Guanine" = "red","Glutamic Acid" = "turquoise1", "Glutamine" = "chocolate1","Glutathione" = "deeppink","liver" = "black")

ImageDimPlot(spotted, group.by = "clusters", cols = metabolite_colors, dark.background = F)
```

The plot highlights the location of each known metabolite. However, due to the inherent issue of mass spectrometry imaging data one m/z mass may be associated with multiple possible metabolites. For example *adenosine triphosphate* or ATP was spiked-in to the second set of wells within the liver (green). When we look at the annotation of ATP, we can see for `m/z-505.988470137` there are 4 different annotations, one of which is ATP. 

```{r, eval=FALSE, warning=FALSE, message=FALSE}
SearchAnnotations(spotted, metabolite = "Adenosine triphosphate", search.exact = T)
```

```{r, echo=FALSE}
df <- SearchAnnotations(spotted, metabolite = "Adenosine triphosphate", search.exact = T)
knitr::kable(df, format = "html", table.attr = "class='dataframe'", align = "c")
```


Plotting this m/z value spatially, it localises completely within the ATP spiked-in region. 

```{r, fig.width=10, fig.height=5, warning=FALSE, message=FALSE}
ImageDimPlot(spotted, group.by = "clusters", cols = setNames(
  ifelse(names(metabolite_colors) == "ATP", "azure", "darkgrey"),names(metabolite_colors)
))+ ggtitle("ATP")| ImageMZPlot(spotted, mzs = 505.9885)
```

This confirms that the correct annotation of `mz-505.988470137` is only 'ATP'. This is the case with all other spiked-in chemicals. To refine the annotations based on pathway correlation we can run SpaMTP's `CalculateAnnotationStatistics`:

```{r, eval=FALSE}
## Step 1: Create a pathway assay - This can be done using only metabolites, or can be merged to include genes if applicable
spotted <- CreatePathwayAssay(spotted, analyte_type = "metabolites", assay = "Spatial", new_assay = "SPM_pathway", verbose = T)

## Step 2: Format Pathway Assay
DefaultAssay(spotted) <- "SPM_pathway"
spotted <- NormalizeData(spotted)
spotted <- ScaleData(spotted)

## Step 3: Run CalculateAnnotationStatistics
output <- CalculateAnnotationStatistics(data = spotted, mz.assay = "Spatial",pathway.assay="SPM_pathway", mz.slot = "counts", return.top = TRUE)
```

After running `CalculateAnnotationStatistics` the output will be a dataframe containing the most likely annotation for each m/z value based on correlations with pathway expression. Lets subset this output to look at our spiked-in metabolites:

```{r, include=FALSE, warning=FALSE, message=FALSE}
annotation_output_file <- resolve_spamtp_vignette_file(
  article = "Annotation_Refinement",
  filename = "annotation_zscore_output.RDS",
  url = "https://zenodo.org/api/records/17352749/files/annotation_zscore_output.RDS/content",
  expected_md5 = "b8d4983f97069c562967880588c3a25e"
)
output <- readRDS(annotation_output_file)
```

```{r, eval=FALSE, warning=FALSE, message=FALSE}
output[c(1162,1053,501,626,126,117,251,113,248,637,638,835,836),]
```

```{r echo=FALSE, results='asis', warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)

df <- output[c(1162,1053,501,626,126,117,251,113,248,637,638,835,836),]
knitr::kable(
  df,
  format = "html",
  table.attr = "style='font-size: 12px; width: 100%;' class='dataframe'",
  align = "c"
)
```


We can see for each spiked-in metabolite the predicted metabolite with the highest significance in each case corresponded to the spike-in. This method gives the user higher confidence that the annotations of m/z value are correct and biologically significant. Note: For large datasets this function can take a while to run, the function `CalculateSingleAnnotationStatistics` can be used instead to calculate refined annotations for a single m/z value. Please visit the [documentation](https://genomicsmachinelearning.github.io/SpaMTP/reference/CalculateSingleAnnotationStatistics.html) for more information.


## 2) Lipid Nomenchlature Refinement 

As described previously in the [`Spatial Metabolomics Analysis`](https://genomicsmachinelearning.github.io/SpaMTP/articles/Mouse_Urinary_Bladder.html) tutorial, because many different kinds of lipids can share the same molecular weight, this may results in some m/z values possessing many possible annotations. The SpaMTP function `RefineLipids` can be used to simplify the lipid nomenclature into common lipid categories and classes, giving more simplified annotations to a level that most SM datasets are capable of identifying. 

```{r, message=FALSE, warning=FALSE, results='hide'}
### Runs lipid nomenclature simplification on annotations
refined_lipid_annotations <- RefineLipids(spotted@assays$Spatial@meta.data, annotation.column = "all_IsomerNames", lipid_info = "simple")
```

We can count the number of metabolites assigned for each m/z before and after refining the lipid annotations.
```{r, message=FALSE, warning=FALSE}
## Count original number of annotations per m/z 
refined_lipid_annotations <- refined_lipid_annotations %>%
  mutate(base_count = sapply(strsplit(all_IsomerNames, ";\\s*"), length))

## Count number of annotations per m/z following lipid simplification
refined_lipid_annotations <- refined_lipid_annotations %>%
  mutate(
    refined_count = ifelse(
      is.na(Lipid.Maps.Category),
      base_count,
      sapply(strsplit(Lipid.Maps.Category, ";\\s*"), length)
    )
  )
```

The maximum reduction in annotations was from 161 to 1. This is quite a significant simplification which can help in processing large datasets. Lets visualise the difference between each m/z value below: 

```{r, fig.width=7, fig.height=7, dpi = 72, warning=FALSE, message=FALSE}
# reshape to long format
df_long <- refined_lipid_annotations %>%
  select(raw_mz, refined_count, base_count) %>%
  pivot_longer(cols = c(refined_count, base_count),
               names_to = "Type", values_to = "Count")

# plot
ggplot(df_long, aes(x = raw_mz, y = Count, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, size = 1.2) +
  scale_color_manual(values = c("refined_count" = "red",
                                "base_count" = "blue"),
                     labels = c("Base annotations", "Refined annotations")) +
  labs(
    x = expression(italic("m/z")),
    y = "Number of annotations",
    color = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.title = element_text(face = "bold")
  )
```

We can see the larger m/z values, which correspond mostly to lipids, display the biggest differences following annotation refinement. 


## 3) Single-Peak Comparison: Custom Matrix-Adduct Design

The standard adduct search space is appropriate only when its ion-formation
rules reflect the experiment. Reactive matrices such as FMP-10 change the
analyte before ion detection. The original FMP-10 workflow represented these
matrix-derived products in the annotation `Adduct` field, which is a useful
computational abstraction even though FMP-10 derivatization is a covalent
reaction followed by possible gas-phase chemistry, rather than a conventional
non-covalent adduct.

### `[M+2FMP10a]+` has a 521.202 Da shift and one charge

The mass of the purchased reagent must not be added directly to the analyte.
FMP-10 is supplied as the iodide salt, `C20H15FIN`, with an average molar mass
of approximately 415.25 g/mol. The iodide is a counter-ion and is not retained
in the detected dopamine derivative. Even the active FMP-10 cation,
`C20H15FN+` (approximately 288.12 Da), is not the net tag mass: during
nucleophilic aromatic substitution, fluorine leaves from FMP-10 and a hydrogen
is lost from the reacting analyte group.

The mass bookkeeping for product I (`2FMP10a`) is therefore:

| Stage | Composition used for the mass change | Approximate monoisotopic change | Charge interpretation |
|:--|:--|--:|:--|
| Purchased FMP-10 iodide salt | `C20H15FIN` | Not used as an adduct shift | Salt contains a cation and iodide counter-ion. |
| One covalently retained tag | `C20H15FN - F - H = C20H14N` | 268.11262 Da | One pyridinium charge is introduced. |
| Two retained tags | `2(C20H14N) = C40H28N2` | 536.22525 Da | The initial double derivative contains two permanent charges. |
| MALDI product I | `C40H28N2 - CH3 = C39H25N2` | 521.202 Da (reported, rounded) | Gas-phase demethylation/charge neutralisation gives the observed singly charged ion. |

Thus, `2FMP10` describes **two derivatization events**, whereas the suffix `a`
identifies one final ion product from those events. It does not mean
`2 * 288.12 Da`, and `z = +1` describes the final ion reaching the detector,
not the charge of the transient double derivative. A second reported product,
`2FMP10b`, follows an alternative dehydrogenation route and has a different net
shift (approximately 535.218 Da). The original study establishes the FMP-10
covalent charge-tagging chemistry and the dopamine signal near `m/z 674.28`;
the subsequent reaction study explicitly describes the singly charged
demethylated and dehydrogenated double derivatives ([original FMP-10
study](https://www.nature.com/articles/s41592-019-0551-3); [reaction-product
study](https://doi.org/10.1002/cmtd.202500062)).

For transparent computation, the rule below derives the product-I shift from
monoisotopic elemental masses. Following the mass-shift convention used in the
reaction study, the elemental composition change and the final charge are
stored separately. `C39H25N2` gives 521.20177 Da, which is reported as
521.202 Da after rounding.

```{r fmp10-mass-bookkeeping}
fmp10_element_mass <- c(
  C = 12,
  H = 1.00782503223,
  N = 14.00307400443
)

# Each substitution retains C20H14N from the active FMP-10 cation.
fmp10_retained_tag_mass <-
  20 * fmp10_element_mass[["C"]] +
  14 * fmp10_element_mass[["H"]] +
  fmp10_element_mass[["N"]]

# Product I: two retained tags, followed by loss of CH3.
# The final +1 charge is recorded separately in the rule table below.
fmp10_double_a_shift <-
  2 * fmp10_retained_tag_mass -
  (fmp10_element_mass[["C"]] + 3 * fmp10_element_mass[["H"]])

fmp10_mass_accounting <- data.frame(
  quantity = c(
    "One retained FMP-10 tag",
    "Two retained tags before demethylation",
    "Final 2FMP10a shift used by SpaMTP"
  ),
  monoisotopic_mass_Da = c(
    fmp10_retained_tag_mass,
    2 * fmp10_retained_tag_mass,
    fmp10_double_a_shift
  )
)

knitr::kable(fmp10_mass_accounting, digits = 7, row.names = FALSE)
```

This exact mass is a **candidate-generation rule**, not proof that a peak is
dopamine. Its credibility comes from restricting the search space with known
experimental chemistry. A defensible FMP-10 search should proceed in the
following order:

1. Search all RaMP compounds with the ordinary positive-mode adduct rules.
2. Add FMP-10 product rules only when the sample was actually prepared with
   FMP-10.
3. Require enough compatible reaction sites for each rule. For `2FMP10a`, the
   candidate should have at least two validated FMP-10-reactive sites
   (phenolic hydroxyl and/or primary or secondary amine sites). Dopamine
   satisfies this requirement.
4. Compare the calculated product m/z with the observed m/z under the chosen
   ppm tolerance, while retaining all mass-compatible alternatives.
5. Award additional confidence for independent evidence such as the expected
   FMP-10 product family, deuterated-FMP-10 shifts, isotope patterns, spatial
   co-localisation, pseudo-MS/MS, or an authentic standard.

If reactive-site counts cannot be obtained from a structure, SpaMTP should
retain the mass match as an exploratory candidate but mark the chemistry as
unverified and lower its prior. In other words, expanding the search space
earns evidential credit only when the acquisition matrix, reaction-site
eligibility, exact mass, and final charge are all recorded; merely adding a
large arbitrary mass shift does not.

This distinction matters for candidate generation. For example, the FMP-10
mouse-brain demonstration contains an observed peak at `m/z 674.2805`, assigned
to double-derivatised dopamine, `[M+2FMP10a]+`. Dopamine has a neutral
monoisotopic mass of approximately 153.07898 Da. None of the standard positive
adduct rules place dopamine near the observed peak:

```{r fmp10-standard-search}
dopamine_peak_mz <- 674.2805

dopamine_db <- chem_props[
  toupper(chem_props$chem_source_id) == "HMDB:HMDB0000073",
  c(
    "ramp_id", "chem_source_id", "inchi_key", "monoisotop_mass",
    "common_name", "mol_formula"
  ),
  drop = FALSE
]
dopamine_db <- dopamine_db[1L, , drop = FALSE]
stopifnot(nrow(dopamine_db) == 1L)

# Dopamine has two phenolic hydroxyls and one primary amine. This explicit
# structure-derived field lets matrix-reaction rules prune impossible products.
dopamine_db$fmp10_reactive_sites <- 3L

standard_positive_rules <- AdductRules("positive")
standard_dopamine_index <- BuildMZAnnotationIndex(
  dopamine_db,
  polarity = "positive",
  rules = standard_positive_rules
)
standard_dopamine_hits <- QueryMZAnnotationIndex(
  dopamine_peak_mz,
  standard_dopamine_index,
  ppm = 5
)

data.frame(
  observed_mz = dopamine_peak_mz,
  search_space = "Standard positive adducts",
  dopamine_matches = nrow(standard_dopamine_hits)
) |>
  knitr::kable(digits = 7, row.names = FALSE)
```

FMP-10 chemistry provides the missing information. Double derivatization can
produce different singly charged products with reported net shifts of
approximately 521.202 and 535.218 Da. These are distinct reaction-product
rules; they should not be generated by blindly multiplying either the salt,
active-cation, or single-tag mass. SpaMTP selects the appropriate product rules
from the declared matrix profile. The user does not need to supply `adducts`;
that argument is only needed to restrict the automatic search space. The matrix
profile extends the standard rules without changing the RaMP compound database:

```{r fmp10-custom-rule}
fmp10_rules <- MALDIMatrixRules(
  maldi_matrix = "FMP-10",
  polarity = "positive"
)

fmp10_product_rules <- fmp10_rules[
  fmp10_rules$rule_class == "reactive_product",
  c(
    "name", "mass_shift", "charge", "reactive_group",
    "min_reactive_sites", "rule_source"
  ),
  drop = FALSE
]
stopifnot(
  isTRUE(all.equal(
    fmp10_product_rules$mass_shift[
      fmp10_product_rules$name == "M+2FMP10a"
    ],
    fmp10_double_a_shift,
    tolerance = 1e-10
  ))
)

knitr::kable(fmp10_product_rules, digits = 7, row.names = FALSE)

# `adducts` is intentionally omitted: use every rule selected by FMP-10.
fmp10_dopamine_index <- BuildMZAnnotationIndex(
  dopamine_db,
  polarity = "positive",
  maldi_matrix = "FMP-10"
)
fmp10_dopamine_hits <- QueryMZAnnotationIndex(
  dopamine_peak_mz,
  fmp10_dopamine_index,
  ppm = 5
)

knitr::kable(
  fmp10_dopamine_hits[, c(
    "observed_mz", "expected_mz", "ppm_error", "adduct", "neutral_mass",
    "metabolite_names", "ramp_ids", "reactive_site_status", "score"
  )],
  digits = 7,
  row.names = FALSE
)
```

Here `mass_shift` and `charge` describe the **net detected ion**. The generic
rule fields enforce the final charge balance expected by the annotation engine;
they do not claim that no hydrogen was lost during the two substitution
reactions. Those reaction-site hydrogen losses, the intermediate `+2` state,
and the subsequent `-CH3` transformation are already folded into the product
rule. Because `fmp10_reactive_sites = 3` was provided, the double derivative is
reported as `verified` for reaction-site eligibility. If this structure-derived
value is unavailable, SpaMTP retains the candidate as `unknown` but applies a
0.25 score multiplier instead of silently treating its chemistry as verified.

The custom rule recovers dopamine at an expected m/z of approximately
674.28075, within 1 ppm of the observed peak. This is a targeted comparison of
the dopamine hypothesis: it does not imply that no other RaMP compound could
match `m/z 674.2805` through a standard or custom rule. A full untargeted
analysis must query the complete RaMP index and report the competing candidates
rather than search dopamine alone.

The rule table is deliberately separate from `chem_props`. A reusable custom
rule should record the reactive matrix, net mass shift, charge, product
variant, polarity, source publication, and a calibrated prior. Apply it only to
experiments using the corresponding chemistry. FMP-10 also supports single and
other multiple-derivatization products, so each validated reaction product
requires its own rule. The FMP-10 reaction and its use for neurotransmitter MSI
are described in the [original
study](https://www.nature.com/articles/s41592-019-0551-3); the alternative
double-derivatization mass shifts are detailed in a later [reaction
study](https://doi.org/10.1002/cmtd.202500062).

This example illustrates the different roles of the evidence layers:

| Evidence layer | Interpretation for `m/z 674.2805` |
|:--|:--|
| Standard adduct search | Dopamine is outside the standard positive-mode search space. |
| Custom FMP-10 rule | The known matrix chemistry makes double-derivatised dopamine a mass-compatible candidate. |
| Pseudo-MS/MS | An independent spectral match can support or challenge that candidate, but should not be fabricated from the adduct rule. |

The existing curated-panel interface remains useful when reference product
masses were measured directly. For example,
`AddFMP10Annotations(..., only.fmp.adduct = TRUE)` applies SpaMTP's curated
FMP-10 product table. The profile/rule approach above is preferable when
designing a transparent, extensible search space for a new reactive matrix or
product variant.

### Automatic matrix profiles and optional adduct selection

The same API covers conventional and reactive MALDI preparation. In the usual
case, the user supplies the experimental matrix and polarity but does not need
to enumerate adducts:

```{r automatic-matrix-api, eval=FALSE}
annotated <- AnnotateSM(
  object,
  assay = "Spatial",
  maldi_matrix = "DHB",
  polarity = "positive",
  ppm_error = 5
)

# Optional restriction for a targeted analysis; not a compulsory argument.
targeted <- AnnotateSM(
  object,
  assay = "Spatial",
  maldi_matrix = "FMP-10",
  polarity = "positive",
  adducts = c("M+FMP10", "M+2FMP10a"),
  ppm_error = 5
)
```

`MALDIMatrixProfiles()` is the auditable registry. It includes widely used
conventional matrices, reactive matrices, and common on-tissue derivatization
reagents, together with the target functional groups and automatic-rule status:

```{r matrix-profile-registry}
matrix_profiles <- MALDIMatrixProfiles()
knitr::kable(
  matrix_profiles[, c(
    "display_name", "category", "default_polarity", "target_groups",
    "automatic_rules"
  )],
  row.names = FALSE
)
```

The profiles deliberately distinguish ionisation behaviour from covalent
reaction chemistry:

| Matrix/profile | Automatic behaviour |
|:--|:--|
| DHB | Uses the standard positive-ion space and adds validated dehydrated-DHB matrix adducts such as `[M+(DHB-H2O)+H]+`. Because DHB also produces abundant matrix clusters, an isolated mass match receives a low prior and should be supported by its parent ion and spatial correlation. |
| CHCA/HCCA | Uses the standard positive-ion space and adds the reported major `[M+CHCA+Na]+` matrix-adduct rule. It is treated as a non-covalent matrix adduct, not a universal covalent tag. |
| 9-AA | Defaults to negative mode and ordinary negative ions, especially `[M-H]-`. No fixed `M+9AA` product is generated because a universal 9-AA metabolite-product shift is not established. |
| FMP-10 | Adds covalent single- and double-derivative products with reaction-site requirements and literature provenance. |
| Other registered reactive matrices/reagents | FMP-8/9, DPP-TFB, TMP-TFB, N-MePyBA, DNPH, coniferyl aldehyde, DHBA, DHAP, Girard T/P, 2-picolylamine, TMPA, AMPP/HATU, and TAHS are represented. A `profile_only` entry selects a sensible polarity and standard ion space but does not invent a universal product mass; a verified study-specific rule can be supplied explicitly. |

For DHB and CHCA matrix adducts, providing `ms1_spectrum` activates the existing
adduct-family check. SpaMTP searches for the corresponding base ion (for
example `[M+H]+` or `[M+Na]+`) in the same spectrum and multiplies an unsupported
matrix-adduct candidate by 0.1. In MSI, the stronger validation is per-pixel
ion-image correlation between the parent and proposed matrix-adduct feature;
that spatial evidence should be stored alongside the annotation when
available.

## 4) Pseudo MS/MS-Based Refinement

### Rationale

Full-scan MSI data do not contain isolated product-ion spectra. However, an
intact ion, its alternative ion forms, and its in/post-source fragments can
have similar spatial distributions. A precursor-specific **pseudo-MS/MS
spectrum** can therefore be reconstructed by grouping m/z features with
correlated ion images and comparing the resulting spectrum with a reference
MS/MS library.

Importantly, the mean spectrum across all pixels is not a precursor-specific
pseudo-MS/MS spectrum. Pooling every detected feature in this way can combine
unrelated metabolites and inflate spectral matches. The workflow used here is
based on the published [MS1-ID
method](https://pmc.ncbi.nlm.nih.gov/articles/PMC11507738/) and its [reference
implementation](https://github.com/Philipbear/ms1_id):

1. Detect and denoise m/z features from the imzML data.
2. Remove features without sufficient spatial structure or pixel coverage.
3. Calculate ion-image correlations and group co-localised features.
4. Construct one pseudo-MS/MS spectrum for each retained target m/z.
5. Search the spectra against unscaled and intensity-scaled GNPS libraries
   using precursor-tolerant reverse spectral matching.
6. Retain the score, matched-peak count, spectral usage, InChIKey, library
   identifier, and all processing parameters as annotation evidence.

Reverse matching ignores additional unmatched peaks in the experimental
pseudo-spectrum, which may originate from correlated background compounds. It
does not make the reconstructed spectrum equivalent to experimentally acquired
MS/MS, and it generally cannot distinguish structural isomers with similar
fragmentation.

### Keep RaMP and pseudo-MS/MS results independent

SpaMTP treats the two resources as separate annotation layers:

| Annotation layer | Primary question | Recommended storage |
|:--|:--|:--|
| RaMP accurate-mass candidates | Which neutral compounds and adducts are compatible with the observed m/z? | The scored SpaMTP annotation table in `object@tools$mz_annotation` |
| MS1-ID/GNPS pseudo-MS/MS evidence | Which library structures are supported by spatially correlated ion forms and fragments? | A separate, versioned result table or RDS database |

The GNPS spectral library is therefore not bundled into the RaMP database and
pseudo-MS/MS scores do not overwrite the RaMP mass, isotope, chemical-validity,
or adduct-network scores. Keeping the results independent also allows either
database to be updated without rebuilding the other. Candidate agreement can
later be assessed using stable identifiers such as InChIKey; compound names
alone should not be used for this comparison.

### Generate the independent pseudo-MS/MS database

MS1-ID is an optional external Python workflow and is not installed or executed
automatically by SpaMTP. The input directory must contain matching `.imzML` and
`.ibd` files. The following example uses both official GNPS library variants:

```{bash pseudo_msms_ms1_id, eval=FALSE}
ms1_id msi \
  --input_dir path/to/msi_data \
  --libs db/gnps_minmz100.pkl db/gnps_minmz100_k10.pkl \
  --mode positive \
  --mz_ppm_tol 5 \
  --min_feature_spatial_chaos 0.10 \
  --min_pixel_overlap 50 \
  --min_correlation 0.85 \
  --lib_search_mztol 0.05 \
  --score_cutoff 0.70 \
  --min_matched_peak 3 \
  --min_spec_usage 0.05 \
  --n_cores 12
```

These values are starting points, not universal confidence boundaries. Mass
accuracy, spatial resolution, the number of tissue pixels, ionisation mode,
and instrument type should guide parameter selection. In particular,
`min_pixel_overlap` must be reduced cautiously for small datasets, and a lower
score threshold increases exploratory coverage at the cost of more false
matches.

For each imzML file, MS1-ID writes `ms1_id_annotations_all.tsv` and a
dereplicated `ms1_id_annotations_derep.tsv`. The complete table should be
retained because it preserves alternative spectral candidates; the
dereplicated table is convenient for inspection but should not be the only
archived result.

```{r import_pseudo_msms, eval=FALSE}
pseudo_msms_file <- file.path(
  "path/to/msi_data",
  "sample_name",
  "ms1_id_annotations_all.tsv"
)

pseudo_msms_results <- read.delim(
  pseudo_msms_file,
  sep = "\t",
  stringsAsFactors = FALSE,
  check.names = FALSE
)

required_columns <- c(
  "pms2_idx", "name", "mz", "spatial_chaos", "matched_score",
  "matched_peak", "spectral_usage", "precursor_mz", "precursor_type",
  "formula", "inchikey", "db_name", "db_id", "pseudo_ms2"
)
stopifnot(all(required_columns %in% colnames(pseudo_msms_results)))
```

Thresholds can be changed by the user when analysing the retained table,
provided that the original MS1-ID search was run with equally permissive or
more permissive cut-offs:

```{r filter_pseudo_msms, eval=FALSE}
pseudo_score_threshold <- 0.70
minimum_matched_peaks <- 3L
minimum_spectral_usage <- 0.05

selected_pseudo_msms <- pseudo_msms_results |>
  dplyr::filter(
    matched_score >= pseudo_score_threshold,
    matched_peak >= minimum_matched_peaks,
    spectral_usage >= minimum_spectral_usage
  ) |>
  dplyr::arrange(mz, dplyr::desc(matched_score),
                 dplyr::desc(matched_peak))
```

For the same dopamine peak, the pseudo-MS/MS table can be queried independently
by observed precursor m/z and InChIKey. The first 14 InChIKey characters compare
the two-dimensional molecular connectivity while avoiding an accidental name
match:

```{r compare_specific_peak_pseudo_msms, eval=FALSE}
dopamine_connectivity_key <- substr(
  toupper(dopamine_db$inchi_key[[1]]),
  1,
  14
)

dopamine_pseudo_msms <- selected_pseudo_msms |>
  dplyr::mutate(
    observed_ppm_error = abs(mz - dopamine_peak_mz) /
      dopamine_peak_mz * 1e6,
    connectivity_key = substr(toupper(inchikey), 1, 14)
  ) |>
  dplyr::filter(
    observed_ppm_error <= 5,
    connectivity_key == dopamine_connectivity_key
  ) |>
  dplyr::arrange(
    dplyr::desc(matched_score),
    dplyr::desc(matched_peak)
  )

dopamine_pseudo_msms[, c(
  "mz", "name", "matched_score", "matched_peak", "spectral_usage",
  "precursor_type", "inchikey", "db_name", "db_id"
)]
```

An empty result means that the independent pseudo-MS/MS layer did not support
dopamine under the selected thresholds; it must not erase the custom-adduct
candidate. Conversely, a spectral match supports the proposed structure but
does not validate the FMP-10 reaction rule by itself. Keeping both rows of
evidence preserves that distinction.

The result and its provenance should be saved together as an independent
database. Record the installed MS1-ID version, GNPS release or checksum, input
file checksum, ionisation mode, and all thresholds used for the analysis.

```{r store_pseudo_msms, eval=FALSE}
pseudo_msms_database <- list(
  metadata = list(
    engine = "ms1_id",
    engine_version = "record the installed version",
    spectral_library = c(
      "gnps_minmz100.pkl",
      "gnps_minmz100_k10.pkl"
    ),
    ion_mode = "positive",
    mz_ppm_tolerance = 5,
    minimum_spatial_correlation = 0.85,
    generated_at = format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
  ),
  results = pseudo_msms_results
)

saveRDS(pseudo_msms_database, "sample_name_pseudo_msms_database.rds")
```

### Interpretation and validation

A pseudo-MS/MS match should be reported as supporting evidence for a candidate,
not as proof of identity. At minimum, inspect the reverse-match score, number
of matched peaks, spectral usage, spatial coherence of the matched ions,
precursor/adduct agreement, and library provenance. Lack of a match is also not
evidence that a RaMP candidate is absent: some metabolites generate few
in-source fragments or are poorly represented in public libraries.

Whenever possible, calibrate the thresholds using authentic standards or
spiked compounds measured on the same platform. Pixel-permuted ion images and
decoy spectral libraries provide useful negative controls. Final reporting
should distinguish accurate-mass candidates, pseudo-MS/MS-supported candidates,
and identifications confirmed by authentic MS/MS or standards.

## 5) Refinement with Paired Targeted Metabolic Data

To come!
