---
title: "SpaMTP: Indexed Metabolite Annotation with RaMP 3.0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SpaMTP: Indexed Metabolite Annotation with RaMP 3.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "##",
  fig.align = "center"
)
options(width = 100)
```

# Overview

SpaMTP's current annotation pipeline separates candidate generation from
downstream filtering. It generates chemically valid adduct hypotheses once,
sorts their expected mass-to-charge ratios, queries that index within a strict
mass tolerance, and ranks the remaining candidates using optional MS1 context.

The workflow is:

1. Select a metabolite database and ion-mode-specific adduct rules.
2. Build a reusable sorted annotation index.
3. Query one or many observed m/z values within a ppm tolerance.
4. Score candidates using mass accuracy, chemical heuristics, isotope evidence,
   and adduct-family support.
5. Store all ppm-valid candidates and choose a score threshold at analysis time.

This article describes upstream candidate annotation. The
[`Annotation Refinement`](Annotation_Refinement.html) article covers downstream
methods for resolving candidates using biological and spatial information.

```{r load-package, message=FALSE, warning=FALSE}
library(SpaMTP)
```

# RaMP snapshot and identifiers

SpaMTP contains a deliberately pruned RaMP snapshot. The annotation table keeps
chemical structures, formulae, monoisotopic masses, names, source identifiers,
and current `RAMP_C_*` identifiers; pathway tables retain only fields required
by SpaMTP's pathway and network functions.

```{r ramp-summary}
ramp_summary <- data.frame(
  property = c(
    "RaMP version",
    "Chemical-property records",
    "Graph nodes after harmonisation",
    "Stale graph IDs remaining"
  ),
  value = c(
    ramp_db_metadata$ramp_version,
    format(ramp_db_metadata$snapshot_counts[["chem_props"]], big.mark = ","),
    format(
      ramp_db_metadata$graph_harmonisation$graph_nodes_after,
      big.mark = ","
    ),
    ramp_db_metadata$graph_harmonisation$stale_ids_remaining
  )
)
knitr::kable(ramp_summary, align = c("l", "r"))
```

The bundled chemical table can be used directly:

```{r chem-props-columns}
data.frame(column = names(chem_props)) |>
  knitr::kable(align = "l")
```

# A small 2-hydroxyglutarate example

The examples below use the RaMP entry for D-2-hydroxyglutaric acid (2HG), with
neutral monoisotopic mass approximately 148.0372 Da. Two carboxyl groups give
this demonstration molecule a maximum of two exchangeable protons.

```{r target-database}
target_db <- chem_props[
  chem_props$ramp_id == "RAMP_C_000219574" &
    grepl("HMDB0000606", chem_props$chem_source_id),
  c(
    "ramp_id", "chem_data_source", "chem_source_id", "inchi_key",
    "monoisotop_mass", "common_name", "mol_formula"
  ),
  drop = FALSE
]
stopifnot(nrow(target_db) > 0L)
target_db <- target_db[1L, , drop = FALSE]
target_db$max_exchangeable_protons <- 2L

knitr::kable(
  target_db[, c(
    "ramp_id", "chem_source_id", "common_name", "mol_formula",
    "monoisotop_mass", "max_exchangeable_protons"
  )],
  digits = 7
)
```

`max_exchangeable_protons` is optional database metadata. When it is present,
SpaMTP removes impossible proton-loss states before indexing. When it is absent,
SpaMTP does not infer this property from molecular formula alone.

# Validated adduct rules

`AdductRules()` returns the ion masses, molecular multiplier, charge,
proton-loss count, chemical class, and prior used by the engine. Bracketed and
unbracketed forms such as `[M+H]+` and `M+H` are both accepted.

```{r adduct-rules}
selected_adducts <- c("M+H", "M+Na", "M+2Na-H", "2M+H")
positive_rules <- AdductRules("positive")
rule_summary <- positive_rules[
  positive_rules$name %in% selected_adducts,
  c(
    "name", "n_molecules", "mass_shift", "charge", "loss_h",
    "complexity", "base_adduct", "prior"
  )
]
rule_summary <- rule_summary[match(selected_adducts, rule_summary$name), ]
knitr::kable(rule_summary, digits = 7, row.names = FALSE)
```

For each rule, SpaMTP calculates

\[
m/z_{expected} =
\frac{nM + \Delta m_{adduct} - \Delta m_{loss}}{|z|}.
\]

The stored `mass_shift` already contains the signed sum of additions and losses.
Custom rule tables must satisfy charge balance before candidate generation.

# Build the reusable index

`BuildMZAnnotationIndex()` performs the expensive expansion once. Expected ion
masses are sorted, allowing `QueryMZAnnotationIndex()` to use binary interval
search rather than scanning every database record for every observed peak.

```{r build-index}
positive_index <- BuildMZAnnotationIndex(
  target_db,
  polarity = "positive",
  adducts = selected_adducts
)
positive_index

ion_table <- data.frame(
  adduct = positive_index$rules$name[positive_index$rule_idx],
  expected_mz = positive_index$expected_mz
)
knitr::kable(ion_table, digits = 9, row.names = FALSE)
```

For large datasets, build one index for each database, polarity, and adduct set,
then reuse it for every spectrum or batch of observed peaks.

## Chemical validity pruning

The 2HG record permits at most two proton losses. Therefore `[M-3H]3-` is not
included in the indexed search space.

```{r proton-pruning}
negative_index <- BuildMZAnnotationIndex(
  target_db,
  polarity = "negative",
  adducts = c("M-H", "M-2H", "M-3H")
)
indexed_negative_rules <- unique(
  negative_index$rules$name[negative_index$rule_idx]
)
data.frame(indexed_rule = indexed_negative_rules) |>
  knitr::kable(align = "l")
```

# Query and rank candidates

The following observed peaks have small controlled mass offsets. The contextual
MS1 spectrum also contains a carbon-13 peak for `[M+H]+` at +1.00335 Da. In
real data, `ms1_spectrum` should represent the same retention-time window,
spatial pixel, or biologically coherent region as the queried peaks.

```{r query-index}
ppm_offsets <- c(1.2, -1.4, 0.6, -0.8)
observed_mz <- ion_table$expected_mz * (1 + ppm_offsets * 1e-6)
names(observed_mz) <- ion_table$adduct

carbon13_spacing <- 1.00335483507
ms1_spectrum <- data.frame(
  mz = c(
    observed_mz,
    ion_table$expected_mz[ion_table$adduct == "M+H"] + carbon13_spacing
  ),
  intensity = c(1000, 700, 300, 200, 53.5)
)

candidates <- QueryMZAnnotationIndex(
  observed_mz,
  positive_index,
  ppm = 5,
  ms1_spectrum = ms1_spectrum,
  min_score = 0
)

candidate_view <- candidates[, c(
  "observed_mz", "expected_mz", "adduct", "ppm_error", "mass_score",
  "rule_prior", "chemical_score", "isotope_score",
  "adduct_network_score", "score", "ramp_ids"
)]
knitr::kable(candidate_view, digits = 4, row.names = FALSE)
```

Candidates outside the requested tolerance are removed before scoring:

```{r ppm-boundary}
outside_five_ppm <-
  ion_table$expected_mz[ion_table$adduct == "M+H"] * (1 + 5.1e-6)

data.frame(
  query = "5.1 ppm outside expected m/z",
  retained_candidates = nrow(QueryMZAnnotationIndex(
    outside_five_ppm,
    positive_index,
    ppm = 5
  ))
) |>
  knitr::kable(row.names = FALSE)
```

The final score is multiplicative:

\[
S = S_{mass} \times P_{rule} \times S_{chemical}
    \times S_{isotope} \times S_{adduct\ family}.
\]

Missing optional MS1 context is neutral rather than evidence of absence. When a
contextual spectrum is supplied, missing expected isotope or base-adduct peaks
can reduce the score.

## Adduct-family support

A complex ion is more credible when a corresponding base monomer is present in
the same spectrum. Here isotope scoring is disabled so that only adduct-network
support changes.

```{r adduct-network}
dimer_mz <- ion_table$expected_mz[ion_table$adduct == "2M+H"]
monomer_mz <- ion_table$expected_mz[ion_table$adduct == "M+H"]

isolated_dimer <- QueryMZAnnotationIndex(
  dimer_mz,
  positive_index,
  ppm = 5,
  ms1_spectrum = data.frame(mz = dimer_mz, intensity = 100),
  check_isotopes = FALSE
)
supported_dimer <- QueryMZAnnotationIndex(
  dimer_mz,
  positive_index,
  ppm = 5,
  ms1_spectrum = data.frame(
    mz = c(dimer_mz, monomer_mz),
    intensity = c(100, 500)
  ),
  check_isotopes = FALSE
)

data.frame(
  context = c("isolated dimer", "base monomer present"),
  adduct_network_score = c(
    isolated_dimer$adduct_network_score,
    supported_dimer$adduct_network_score
  ),
  final_score = c(isolated_dimer$score, supported_dimer$score)
) |>
  knitr::kable(digits = 3, row.names = FALSE)
```

## Mass-defect heuristic

For a nominally CHO-only candidate, an unusually negative signed mass defect is
treated as background-like evidence. It receives a penalty rather than being
deleted, because halogenated, metal-containing, and unusual biological ions
must not be rejected by a universal fractional-mass rule.

```{r mass-defect}
background_like_db <- data.frame(
  formula = "C2H2O2",
  exactmass = 99.942723533379,
  id = "synthetic-background-like-candidate",
  name = "Synthetic background-like candidate"
)
background_index <- BuildMZAnnotationIndex(
  background_like_db,
  polarity = "positive",
  adducts = "M+H"
)
background_mz <- background_index$expected_mz[[1L]]

with_penalty <- QueryMZAnnotationIndex(
  background_mz,
  background_index,
  ppm = 0,
  use_mass_defect = TRUE,
  check_isotopes = FALSE,
  check_adduct_network = FALSE
)
without_penalty <- QueryMZAnnotationIndex(
  background_mz,
  background_index,
  ppm = 0,
  use_mass_defect = FALSE,
  check_isotopes = FALSE,
  check_adduct_network = FALSE
)

data.frame(
  setting = c("mass-defect heuristic on", "mass-defect heuristic off"),
  signed_mass_defect = background_mz - round(background_mz),
  chemical_score = c(
    with_penalty$chemical_score,
    without_penalty$chemical_score
  ),
  retained = TRUE
) |>
  knitr::kable(digits = 3, row.names = FALSE)
```

# Annotate individual values or a SpaMTP object

`AnnotateMZ()` is a convenience wrapper for individual or small batches of
observed values. Supply a pre-built index to avoid rebuilding it.

```{r annotate-mz}
single_result <- AnnotateMZ(
  observed_mz[["M+H"]],
  index = positive_index,
  polarity = "positive",
  ppm = 5,
  ms1_spectrum = ms1_spectrum,
  min_score = 0
)
knitr::kable(
  single_result[, c(
    "observed_mz", "metabolite_names", "ramp_ids", "adduct",
    "ppm_error", "score"
  )],
  digits = 4,
  row.names = FALSE
)
```

`AnnotateSM()` applies the same engine to m/z feature metadata and preserves the
ranked candidate table in `@tools$mz_annotation`. The small object below is only
for demonstrating storage and provenance.

```{r annotate-object}
demo_counts <- Matrix::Matrix(
  matrix(
    seq_len(length(observed_mz) * 4L),
    nrow = length(observed_mz),
    ncol = 4L
  ),
  sparse = TRUE
)
rownames(demo_counts) <- paste0("mz-", observed_mz)
colnames(demo_counts) <- paste0("pixel_", seq_len(ncol(demo_counts)))

demo_object <- SeuratObject::CreateSeuratObject(
  counts = demo_counts,
  assay = "SPM"
)
feature_metadata <- data.frame(
  raw_mz = unname(observed_mz),
  row.names = rownames(demo_object[["SPM"]])
)
demo_object[["SPM"]] <- SeuratObject::AddMetaData(
  demo_object[["SPM"]],
  metadata = feature_metadata
)

annotated_object <- AnnotateSM(
  demo_object,
  db = NULL,
  assay = "SPM",
  raw.mz.column = "raw_mz",
  ppm_error = 5,
  adducts = selected_adducts,
  polarity = "positive",
  return.only.annotated = FALSE,
  save.intermediate = TRUE,
  min_score = 0,
  verbose = FALSE,
  index = positive_index,
  ms1_spectrum = ms1_spectrum
)

annotation_columns <- annotated_object[["SPM"]][[]][, c(
  "raw_mz", "all_IsomerNames", "all_Adducts", "all_Scores",
  "all_Ramp_IDs"
)]
knitr::kable(annotation_columns, digits = 4)
```

`AnnotationInfo()` records which engine and database provenance downstream
functions will consume.

```{r annotation-info}
annotation_info <- AnnotationInfo(annotated_object)
annotation_info_table <- data.frame(
  field = names(annotation_info),
  value = vapply(annotation_info, paste, collapse = "; ", character(1)),
  row.names = NULL
)
knitr::kable(annotation_info_table, row.names = FALSE)
```

# Choose thresholds without re-annotation

The recommended production pattern is to retain all ppm-valid candidates using
`min_score = 0`, then select a threshold appropriate to the analysis. This makes
coverage and specificity tunable without repeating index construction.

```{r score-thresholds}
stored_candidates <- annotated_object@tools$mz_annotation$results
score_thresholds <- c(0, 0.2, 0.5)
threshold_summary <- data.frame(
  score_threshold = score_thresholds,
  retained_candidates = vapply(
    score_thresholds,
    function(threshold) sum(stored_candidates$Score >= threshold),
    integer(1)
  )
)
knitr::kable(threshold_summary, row.names = FALSE)
```

For pathway analysis, the annotation threshold and differential-significance
threshold answer different questions. Setting `pval_cutoff_mets = 1` allows
non-DE metabolites to contribute, while `annotation_score_threshold` still
controls which chemical candidates are eligible.

```{r downstream-thresholds, eval=FALSE}
regional_pathways <- FindRegionalPathways(
  object,
  analyte_types = c("genes", "metabolites"),
  ident = "region",
  DE.list = differential_results,
  pval_cutoff_mets = 1,
  annotation_score_threshold = 0.05
)

PathwayNetworkPlots(
  object,
  ident = "region",
  regpathway = regional_pathways,
  DE.list = differential_results,
  annotation_score_threshold = 0.05,
  metabolite_detection = "annotated"
)
```

# Production use with the complete RaMP table

For one call, `AnnotateSM(db = NULL)` automatically uses the bundled RaMP
chemical-property table. For repeated calls, explicitly build and reuse an
index. Restricting the adduct set to ions supported by the experiment reduces
memory use and false hypotheses.

```{r production-pattern, eval=FALSE}
ramp_positive_index <- BuildMZAnnotationIndex(
  chem_props,
  polarity = "positive",
  adducts = c("M+H", "M+Na", "M+K")
)

object <- AnnotateSM(
  object,
  db = NULL,
  index = ramp_positive_index,
  assay = "SPM",
  polarity = "positive",
  ppm_error = 5,
  adducts = c("M+H", "M+Na", "M+K"),
  min_score = 0,
  return.only.annotated = FALSE
)
```

# Interpretation and limitations

- An annotation is a ranked chemical hypothesis, not a confirmed identity,
  probability, or false-discovery rate.
- Accurate mass alone generally cannot resolve structural isomers.
- Isotope and adduct-family checks require a spectrum from an appropriate
  retention-time, pixel, or regional context.
- The mass-defect rule is a score penalty, not a universal exclusion rule.
- Exchangeable-proton pruning requires the relevant database property; it is
  not inferred from elemental formula alone.
- Orthogonal evidence such as authentic standards, retention time, ion
  mobility, or MS/MS remains necessary for high-confidence identification.

The [`Annotation Refinement`](Annotation_Refinement.html) workflow can then use
spatial, pathway, lipid-class, or pseudo-MS/MS information to further prioritise
the candidates stored here.

```{r session-info}
sessionInfo()
```
