---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SpaMTP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 150
)
```

```{r fxns, include = FALSE}
### Code Modified from Seurat: https://github.com/satijalab/seurat/
library('htmlTable')
make_list <- function(items) {
  paste0("<ul>", sprintf('<li>%s</li>', items), '</ul>', collapse = '')
}
make_href <- function(url, text){
  paste0("<div style='text-align:center; font-size:large; font-weight:bold'><a href=", url ,">", text, "</a></div>")
}
make_href2 <- function(url, text){
  paste0("<a href=", url ,">", text, "</a>")
}
process_entry <- function(dat) {
  if (grepl(pattern = "https://satijalab.org/img/vignette_images", x = dat$image)) {
    img <- paste0('![](', dat$image, '){width=3000px}')
  } else if (grepl(pattern = "assets/", x= dat$image)) {
    img <- paste0('![](', dat$image, '){width=3000px}') 
  } else {
    img <- paste0('![](', '../output/images/', dat$image, '){width=3000px}')
  }
  if (grepl(pattern = "https://satijalab.org/", x = dat$name)) {
    link <- dat$name
  } else {
    link <- paste0(dat$name, ".html")
  }
  go.button <- paste0('<a class="btn btn-primary" href="', link, '" role="button" style="width:100%">GO</a>')
  data.frame(
    title = make_href(url = link, text = dat$title),
    img = img,
    desc = dat$summary,
    btn = go.button
  )
}
process_wrapper_entry <- function(dat) {
  data.frame(
    Package = dat$name,
    Vignette = make_href2(url = dat$link, text = dat$title),
    Reference = make_href2(url = dat$reference, text = dat$citation),
    Source = make_href2(url = dat$source, text = dat$source)
  )
}
make_vignette_card_section <- function(vdat) {
  vignettes <- vdat$vignettes
  dat <- data.frame(title = character(), img = character(), desc = character())
  for (v in 1:length(x = vignettes)) {
    dat <- rbind(dat, process_entry(vignettes[[v]]))
    if(nrow(x = dat) == 3 | v == length(x = vignettes)){
      colnames(dat) <- NULL
      dat <- t(dat)
      if (ncol(x = dat) == 2) {
        print(htmlTable(
          dat, 
          align = '|l|l|', 
          css.cell = "padding-left: .75em; width: 50%", 
          css.class = "two-column-htmltable"
        ))
      } else if (ncol(x = dat) == 1){
        print(htmlTable(
          dat, 
          align = '|l|', 
          css.cell = "padding-left: .75em; width: 100%", 
          css.class = "one-column-htmltable"
        ))
      } else {
        print(htmlTable(
          dat, 
          align = '|l|l|l|', 
          css.cell = "padding-left: .75em; width: 30%; background-color: white; border-color: rgba(0,0,0,0); color: black;"
        ))
      }
      dat <- data.frame(title = character(), img = character(), desc = character())
    }
  }
}
```

```{r yaml, include = FALSE}
library(yaml)
vdat <- read_yaml(file = "get_started.yaml")
```

```{=html}
<style>
.two-column-htmltable {
  width: 66.666%;
  background-color: white;
}
.one-column-htmltable {
  width: 33.333%;
  background-color: white;
}

</style>
```

Listed below are a number of useful tutorials demonstrating how to use SpaMTP when analysing your spatial metabolomic datasets. For more documentation of each function used in these tutorials, please visit our [Reference page](../reference/index.html).


## Useful SpaMTP Vignettes

For a general introduction in importing data, annotating metabolites and running key analysis pipelines on your spatial metabolomic data, we suggest starting with the Spatial Metabolomic Analysis Tutorial. This vignette uses mouse bladder data and demonstrates how to perform general tasks such as plotting, subsetting, and manipulating *SpaMTP* Seurat objects. For those with paired multi-modal datasets, we suggest working through the Multi-Modal Integration Tutorial. Here, we provide methods to align your two datasets to the same coordinates, map spatial metabolomic pixels to spatial transcriptomic spots and then run various analysis using this integrated data. 

```{r results='asis', echo=FALSE, warning=FALSE, message = FALSE, dev.args = list(bg = "#ffffff")}
make_vignette_card_section(vdat = vdat)
```

## SpaMTP Public Datasets

To access and download all our datasets used in the vignettes you can vist the [*SpaMTP* zenodo page](https://zenodo.org/communities/spamtp/records).  


## Installation

You can install the current version of SpaMTP from
[GitHub](https://github.com/) with:

``` r
if (!require("devtools", quietly = TRUE))
    install.packages("devtools")

devtools::install_github("GenomicsMachineLearning/SpaMTP")
```
<br>

### Installing with *conda* environment

Below is an example of how to set up SpaMTP using a conda environment:

``` console
conda create -n SpaMTP -c conda-forge r-base=4.3.3 r-essentials r-devtools r-cairo python=3.9 r-reticulate r-leiden r-imager r-magick r-hdf5r r-sf bioconda::bioconductor-ebimage bioconda::bioconductor-scater bioconda::bioconductor-dropletutils bioconda::bioconductor-fgsea bioconda::bioconductor-rgoslin bioconda::bioconductor-glmgampoi
conda activate SpaMTP
```

Within the SpaMTP environment open *R* and install ***SpaMTP***.

``` r
#Install Seurat version 5.3.0
remotes::install_github("satijalab/seurat", "seurat5", quiet = FALSE)

#Install Cardinal
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("Cardinal")

#Install SpaMTP
devtools::install_github("GenomicsMachineLearning/SpaMTP")
```

<br>

### Possible Installiation Errors

#### Cairo Failed to Install

``` console
checking if R was compiled with the RConn patch... no
checking for cairo.h... no
configure: error: Cannot find cairo.h! Please install cairo (http://www.cairographics.org/) and/or set CAIRO_CFLAGS/LIBS correspondingly.
ERROR: configuration failed for package ‘Cairo’
```

Check these libraries exist:

``` console
conda install conda-forge::r-cairo
```

This should resolve any issues. Rerun:

``` r
BiocManager::install("scater")
```

#### Cardinal Failed to Install

``` console
ERROR: dependency ‘EBImage’ is not available for package ‘Cardinal’

BiocManager::install("EBImage")

fftwtools.c:28:9: fatal error: fftw3.h: No such file or directory
   28 | #include<fftw3.h>
      |         ^~~~~~~~~
```

If EBImage failed to installed, it is most likely due to an issue with the installation of *fftwtools*.

Try:

``` console
conda config --add channels conda-forge
conda config --set channel_priority strict

conda install r-fftwtools
```

Else try:

``` console
conda install bioconda::r-fftwtools
conda install conda-forge::fftw
```

Try in *R*:

```r

if (!require("BiocManager", quietly = TRUE)) #Check if BiocManager is installed
    install.packages("BiocManager")

BiocManager::install("fftwtools")

```

If none of the above methods resolve instiallation issues, install *Cardinal* directly through *conda*:

``` console
conda install bioconda::bioconductor-cardinal  
```

#### *rgoslin* Failed to Install

If rgoslin fails to install please head to their [github](https://github.com/lifs-tools/rgoslin). Alternative, you can try installing through BiocManager or mamba shown below:

```r
BiocManager::install("rgoslin")
```
or
```console
mamba install bioconductor-rgoslin
```


#### *Matrix* version incompatible with *Seurat*

If the *Matrix* package version installed is not compatible with *Seurat* then try installing a different version and re-installing *Seurat*

```r
#Install Correct version of Matrix package
remotes::install_version("Matrix", version = "1.6-4", repos = "https://cran.r-project.org")

#Install Seurat version 5.1.0
install.packages("Seruat")
```

For other issue please flag on github under *Issues*.



