---
title: "SpaMTP: Spatial Metabolite Annotation"
author: "Tianyao Lu"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{SpaMTP: Spatial Metabolite Annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.align = "center"
)
```

# Overview

`SpaMTP` provides analysis and visualisation methods for spatial metabolomics
and paired spatial multi-omics experiments. It uses standard Seurat, Cardinal,
and Bioconductor infrastructure while keeping versioned annotation resources in
the companion `SpaMTPdb` package and experiment objects in `SpaMTPData`.

This vignette demonstrates a small, fully reproducible accurate-mass annotation
workflow. It deliberately uses an in-memory database so that package checking
does not require a network connection or a large data download.

```{r load-package, message=FALSE}
library(SpaMTP)
```

# Inspect available database resources

`SpaMTPDatabaseInfo()` reads the lightweight SpaMTPdb registry without
downloading any annotation table.

```{r database-registry}
database_registry <- SpaMTPDatabaseInfo(version = "latest")
database_registry[
  database_registry$resource %in% c("chem_props", "smiles_features"),
  c("resource", "version", "category", "rows", "columns")
]
```

# Create a small structure-aware database

The example contains D-2-hydroxyglutaric acid and glucose. SMILES connectivity
is decomposed into interpretable functional groups and ionisation sites before
candidate generation.

```{r demo-database}
demo_database <- data.frame(
  ramp_id = c("RAMP_C_2HG", "RAMP_C_GLUCOSE"),
  chem_source_id = c("HMDB:HMDB0000606", "HMDB:HMDB0000122"),
  common_name = c("D-2-hydroxyglutaric acid", "D-glucose"),
  monoisotop_mass = c(148.037173, 180.063388),
  mol_formula = c("C5H8O5", "C6H12O6"),
  iso_smiles = c("O=C(O)CC(O)C(=O)O", "OCC1OC(O)C(O)C(O)C1O"),
  stringsAsFactors = FALSE
)

demo_database <- AnnotateSMILESStructure(demo_database)
demo_database[, c(
  "common_name", "carboxyl_sites", "hydroxyl_sites",
  "max_exchangeable_protons", "positive_mode_score",
  "negative_mode_score", "alkali_affinity_score"
)]
```

# Build and query an indexed search space

Candidate masses are generated once and sorted. Repeated observed peaks can
then be queried using a binary range search rather than scanning the full
database for every peak.

```{r indexed-search}
annotation_index <- BuildMZAnnotationIndex(
  db = demo_database,
  polarity = "negative",
  adducts = c("[M-H]-", "[2M-H]-"),
  infer_structure = "always"
)

observed_mz <- 147.02935
candidates <- QueryMZAnnotationIndex(
  observed_mz = observed_mz,
  index = annotation_index,
  ppm = 5
)

candidates[, c(
  "observed_mz", "expected_mz", "metabolite_names", "adduct",
  "ppm_error", "score"
)]
```

# Reproducibility

For production analyses, record the database version returned by
`SpaMTPDatabaseInfo()` together with ion mode, matrix profile, adduct rules,
mass tolerance, and annotation-score threshold.

```{r session-info}
sessionInfo()
```
